\Large \bf Cyclone: High Availability for Persistent Key Value Stores

### 2017-07-29 paper/paper.tex
Updates.

These key value stores however tend to ignore an important component: the write
ahead log. A machine or storage failure leading to total loss or temporary
unavailability of data is unacceptable in services where high availabilty and
revenue are interconnected. Key value stores therefore usually incorporate
support for a write ahead log that if replicated and kept durable for every
appended update provides the necessary high availability. Unfortunately, the
write ahead log is a performance achilles heel for these systems - eclipsing
much of the work on improving the performance of the LSM component. To
illustrate the impact of the write ahead log, consider
Figure~\ref{fig:problem}. The line marked 'Rocksdb' shows the performance of
Rocksdb without the write ahead log. The performance when persisting the write
ahead log without replicating it is shown as the line marked 'rocksdb/WAL'. The
line marked 'rocksdb/3 way rep.' is for simply replicating the log three ways
without persisting it, using RAFT~\cite{raft} running over TCP/IP. Either
persisting every update to the log or replicating it using TCP/IP causes
performance to drop by an order of magnitude (note the log scale on the
x-axis). It is therefore no surprise that deployments of Rocksdb often turn off
the write ahead log~\cite{samza}. On the other side of the spectrum, key value
store research prototypes such as FloDB~\cite{flodb} turns off the write ahead
log to be able to showcase benefits of sophisticated extensions to LSM
datastructures.

@@

The first key contribution of Cyclone is demonstrating the benefit of a small
amount of non-volatile memory when combined with the flash SSD based persistent
logs.  The primary cause of slowdown when turning on a write ahead log on a
single machine in Figure~\ref{fig:problem} is the need to do synchronous block
IO for every update \emph{regardless of the size of the update}. SSDs provide
the maximum throughput when writing large chunks (usually more than 4KB) of
data. For small updates, the synchronous write latency dominates the time for
IO.  Cyclone avoids this by making use of directly attached persistent memory on
the server. This can be in the form of NVDIMMs (DIMMs with an added
ultracapacitor to dump state to a small amount of attached flash) or newer forms
of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs are expensive
- retailing at approximately 2X the price of a standard DIMM [find ref]. This
means that they will likely be available in small quantities. We compensate for
this by periodically draining the NVDIMM log into a log placed on a standard SSD
on the system.

@@

The rest of the paper is organized as follows. We describe Cyclone's system
architecture in Section~\ref{sec:sysarch}. We describe this two level log
structure in Section~\ref{sec:storage}. The set of optimizations to the network
stack is decribed in Section~\ref{sec:network}. Both the storage (NVM, SSD) as
well as the NIC expose significant parallelism that is left unused when Cyclone
is used to replicate a single log. To exploit this parallelism we use Cyclone to
replicate multiple logs, each using a different instance of the RAFT consensus
protocol. We show how to exploit the commutativity inherent in key-value store
interfaces to make use of this parallelism in Cyclone's design in
Section~\ref{sec:parallelism}.  A detailed evaluation of Cyclone's replication
is provided in the context of RocksDB - a popular persistent key value store -
in Section~\ref{sec:evaluation}. We then discuss related work before concluding.

@@

Cyclone integrates with key value stores as both a client and server side
library. Cyclone replicates the key value store across a set of replica
servers. The client side library sends requests to a distinguished leader
replica.  On the server side Cyclone accepts requests from clients and calls
into the key value store. It then returns a response to the client. Before
executing any PUT request, Cyclone appends it to a durable log and replicates
the request to the logs of follower replicas. The request is considered
replicated once acknowledged by a majority of replicas.

Cyclone uses RAFT~\cite{raft} to keep the logs in sync across the replicas. On a
failure of the leader replica a new leader is automatically elected - RAFT
ensures that the new leader has the most up to date logs. The Cyclone client
library automatically locates the new leader on a failover - clients therefore
see only a brief interruption in service. Although read requests (GET
operations) are not logged, they are serviced in FIFO order with the write requests
(PUT operations) currently in Cyclone's log. This ensures linearizability in the
face of failovers.

@@

Cyclone adopts a different strategy by making use of novel memory technologies
to obviate the need to make this tradeoff in the first place. We assume a small
amount of non-volatile memory directly addressable by the CPU rather than being placed
behind a block oriented IO interface. Log appends are done to the ``NVM log''
placed in this non-volatile memory. While we designed Cyclone to leverage new
non-volatile memory technologies that might be available in the
future~\cite{3dxpoint} current non-volatile memory options such are NVDIMMs
are more expensive that DRAM. Therefore, for Cyclone to be economical
we cannot assume that a sufficient amount of NVM is available to hold the entire
log. The log entries can only be garbage collected when the in-memory component
of the key value store is checkpointed to secondary storage. The growing size of
available DRAM means that also provisioning sufficient NVDIMMs to hold the
corresponding log of updates is not feasible given that NVDIMMs cost at least
twice as much as volatile DRAM. We therefore resort to using a second larger log
level placed on a flash SSD. Entries are drained from the NVM log to this flash
log in conveniently large units that are a multiple of 4KB - the optimum IO unit
for flash SSDs. This ensures we get the best
possible throughput from a flash device without paying the price of synchronous
IO for small key-value pairs. At the same time, the amount of non-volatile
memory required is small enough to not add undue cost of the server.

@@

In order to recover from a crash, we make two important assumptions about the
underlying SSD. First, we assume that 4KB is the \emph{minimum} atomic unit for
updating pages on the SSD even under power failure i.e. there are no shorn
writes (otherwise known as torn writes) on a 4KB page~\cite{shorn_writes}. We
also assume that the SSD has power loss data protection meaning that writes
cached in the drive's volatile cache are written to the SSD using a backup
capacitor in the event of power failure - a property of many data center class
SSDs today, including the Intel DC P3600 SSD~\cite{ssd_spec} we use in our
evaluation. Together these two assumptions mean that we can recover a consistent
prefix of the log on a power failure.

@@

The two level log design in Cyclone lays the foundation for improvements to the
network stack that we discuss next - a significant contributor to the overheads
in Figure~\ref{fig:problem}. We use RAFT to replicate the NVM log from the
leader replica to the followers. Each replica \emph{independently} drains
committed entries (in terms of RAFT consensus) from the NVM log to the
flashlog. Block IO therefore is no longer a consideration when optimizing data
movement over the network, unlike systems that need to deal with such
problems~\cite{reflex}. Replication reduces to efficiently sending a block of
data already present in directly attached memory over the network.

@@

\subsection{Eliminating Data Movement}
Software packet switches built on DPDK try to touch as little of the packet data
as possible, minimizing movement of data up and down the cache and memory
hierarchy. Log replication looks like multicast that does not require deep
packet inspection. Software packet switches implementing multicast therefore
simply manipulate packet headers to produce new packets to send on the output
ports. We designed our implementation of raft around the same principle.

@@

@@

@@

The scalable commutativity rule~\cite{scalable_commutativity} generalizes the
idea of commutativity in key value store interfaces that we have used for
horizontal scaling in Cyclone. Conveniently, it provides a tool for checking when
API calls can commute in a history. It is a simple exercise to express a key
value store API in COMMUTER and check that operations to different keys
commute. This provides a theoretical basis for horizontal scaling, something we
have omitted for brevity from this paper. The scalable commutativity rule also
suggests that scalable key value store implementations would necessarily have
commutative interfaces, something we confirm and exploit for key
value stores.

>>>>

These key value stores however tend to ignore an important component: the write
ahead log. A machine or storage failure leading to total loss or temporary
unavailability of data is unacceptable in services where high availabilty and
revenue are interconnected. Key value stores therefore usually incorporate
support for a write ahead log that if replicated and kept durable for every
appended update provides the necessary high availability. Unfortunately, the
write ahead log is a performance achilles heel for these systems - eclipsing
much of the work on improving the performance of the LSM component. To
illustrate the impact of the write ahead log, consider
Figure~\ref{fig:problem}. The line marked 'Rocksdb' shows the performance of
Rocksdb without the write ahead log. The performance when persisting the write
ahead log without replicating it is shown as the line marked 'rocksdb/WAL'. The
line marked 'rocksdb/3 way rep.' is for simply replicating the log three ways
without persisting it, using RAFT~\cite{raft} running over TCP/IP. Either
persisting every update to the log or replicating it using TCP/IP causes
performance to drop by an order of magnitude (note the log scale on the
x-axis). It is therefore no surprise that deployments of Rocksdb often turn off
the write ahead log~\cite{samza}, depending on upper layers to provide
availability by co-ordinating replicas. On the other side of the spectrum, key
value store research prototypes such as FloDB~\cite{flodb} turn off the write
ahead log to be able to showcase benefits of sophisticated extensions to LSM
datastructures.

@@

The first key contribution of Cyclone is demonstrating the benefit of a small
amount of non-volatile memory when combined with the flash SSD based persistent
logs.  The primary cause of slowdown when turning on a write ahead log on a
single machine in Figure~\ref{fig:problem} is the need to do synchronous block
IO for every update \emph{regardless of the size of the update}. SSDs provide
the maximum throughput when writing large chunks (usually more than 4KB) of
data. This is due to the fixed cost of a round trip to the SSD through the IO
interface (NVMe in our case), regardless of the volume of data being
synchronously persisted. Cyclone avoids this by making use of directly attached
persistent memory on the server. This can be in the form of NVDIMMs (DIMMs with
an added ultracapacitor to dump state to a small amount of attached flash) or
newer forms of persistent memory~\cite{pmfs}. For current deployments, NVDIMMs
are expensive - retailing at approximately 2X the price of a standard DIMM [find
  ref]. This means that they will likely be available in small quantities. We
compensate for this by periodically draining the NVDIMM log into a log placed on
a standard SSD on the system.

@@

The rest of the paper is organized as follows. We describe Cyclone's system
architecture in Section~\ref{sec:sysarch}. We describe this two level log
structure in Section~\ref{sec:storage}. The set of optimizations to the network
stack is decribed in Section~\ref{sec:network}. Both the storage (NVM, SSD) as
well as the NIC expose significant parallelism that is left unused when Cyclone
is used to replicate a single log. To exploit this parallelism we use Cyclone to
replicate multiple logs, each using a different instance of the RAFT consensus
protocol. We show how to exploit the commutativity inherent in key-value store
interfaces to make use of this parallelism in Cyclone's design in
Section~\ref{sec:parallelism}.  A detailed evaluation of Cyclone's replication
performance is provided in the context of RocksDB - a popular persistent key
value store - in Section~\ref{sec:evaluation}. We then discuss related work
before concluding.

@@

Cyclone integrates with key value stores as both a client and server side
library. Cyclone replicates the key value store across a set of replica
servers. The client side library sends \emph{all} requests to a distinguished
leader replica.  On the server side Cyclone accepts requests from clients and calls
into the key value store. It then returns a response to the client. Before
executing any PUT request, Cyclone appends it to a durable log and replicates
the request to the logs of follower replicas. The request is considered
replicated once acknowledged by a majority of replicas. Follower replicas apply
updates from the log in order.

Cyclone uses RAFT~\cite{raft} to keep the logs in sync across the replicas. On a
failure of the leader replica a new leader is automatically elected - RAFT
ensures that the new leader has the most up to date log. The Cyclone client
library automatically locates the new leader on a failover - clients therefore
see only a brief interruption in service. Although read requests (GET
operations) are not logged, they are serviced in FIFO order with the write requests
(PUT operations) currently in Cyclone's log at the moment the read request is
received. This ensures linearizability in the face of failovers.

@@

Cyclone adopts a different strategy by making use of novel memory technologies
to obviate the need to make this tradeoff in the first place. We assume a small
amount of non-volatile memory directly addressable by the CPU rather than being
placed behind a block oriented IO interface. Log appends are done to the ``NVM
log'' placed in this non-volatile memory. While we designed Cyclone to leverage
new non-volatile memory technologies that might be available in the
future~\cite{3dxpoint}, current non-volatile memory options such are NVDIMMs are
more expensive that DRAM. Therefore, for Cyclone to be economical we cannot
assume that a sufficient amount of NVM is available to hold the entire log. The
log entries can only be garbage collected when the in-memory component of the
key value store is checkpointed to secondary storage. The growing size of
available DRAM means that also provisioning sufficient NVDIMMs to hold the
corresponding log of updates is not feasible given that NVDIMMs cost at least
twice as much as volatile DRAM. We therefore resort to using a second larger log
level placed on a flash SSD. Entries are drained from the NVM log to this flash
log in conveniently large units that are a multiple of 4KB - the optimum IO unit
for flash SSDs. This ensures we get the best possible throughput from a flash
device without paying the price of synchronous IO for small key-value pairs. At
the same time, the amount of non-volatile memory required is small enough to not
add undue cost of the server.

@@

In order to recover from a crash, we make two important assumptions about the
underlying SSD. First, we assume that 4KB is the \emph{minimum} atomic unit for
updating pages on the SSD even under power failure i.e. there are no shorn
writes (otherwise known as torn writes) on a 4KB page~\cite{shorn_writes}. We
also assume that the SSD has power loss data protection meaning that writes
cached in the drive's volatile cache are written to the SSD using a backup
capacitor in the event of power failure - a property of many data center class
SSDs today, including the Intel DC P3600 SSD~\cite{ssd_spec} we use in our
evaluation. Together, these two assumptions mean that we can recover a consistent
prefix of the log on a power failure.

@@

The two level log design in Cyclone lays the foundation for improvements to the
network stack that we discuss next - a significant contributor to the overheads
in Figure~\ref{fig:problem}. We use RAFT to replicate the NVM log from the
leader replica to the followers. Each replica \emph{independently} drains
committed (in terms of RAFT consensus) entries from the NVM log to the
flashlog. Block IO therefore is no longer a consideration when optimizing data
movement over the network, unlike systems that need to deal with such
problems~\cite{reflex}. Replication reduces to efficiently sending a block of
data already present in directly attached memory over the network.

@@

\subsection{Eliminating Data Movement}
\label{sec:dm}
Software packet switches built on DPDK try to touch as little of the packet data
as possible, minimizing movement of data up and down the cache and memory
hierarchy. Log replication looks like multicast that does not require deep
packet inspection. Software packet switches implementing multicast therefore
simply manipulate packet headers to produce new packets to send on the output
ports. We designed our implementation of raft around the same principle.

@@

Finally, we discuss our choice of RAFT as a consensus protocol. A key advantage
of RAFT is that unlike alternatives such as MultiPaxos~\cite{multipaxos} or
Quorum based replication~\cite{quorum}, the leader is guaranteed to have the
most up to date logs. There is therefore no case where a leader needs to receive
log entries to ``fill'' holes in its logs from follower replicas, simplifying
the protocol state machine we need to implement. A leader can immediately reject
any responses that are not more current than its view (term). In turn, the
persistent log operates as a double ended queue, with entries either being
appended (or possibly deleted at followers) at the end and being deleted from
the front (after commit). This is critical to efficient operation of Cyclone, as
it allows a simple top level structure (a circular buffer of pointers) to
represent the log.

@@

DARE~\cite{dare} implements state machine replication over RDMA, using a custom
replication protocol. It is worthwhile to point out that since DARE allows any
replica to become the leader it requires a log adjustment protocol for the
leader to bring its logs up to date, a situation we avoid by using RAFT as
described in Section~\ref{sec:dm}. Using 40 Gb/s speeds, and 3-way replication
DARE reported being able to replicate 500K requests/sec. This is without
persisting the log of updates. In contrast, Cyclone replicates upwards of 6
million similarly sized requests each second with 4*10 Gbps ethernet (without the
second level flashlog). The conclusions we can draw are similar to the
comparison with Mojim: a carefully implemented log replication system obviates
the need for network offload.

@@

The scalable commutativity rule~\cite{scalable_commutativity} generalizes the
idea of commutativity in key value store interfaces that we have used for
horizontal scaling in Cyclone. Conveniently, it provides a tool for checking when
API calls can commute in a history. It is a simple exercise to express a key
value store API in COMMUTER and check that operations to different keys
commute. This provides a theoretical basis for horizontal scaling, something we
have omitted for brevity from this paper. The scalable commutativity rule also
suggests that scalable key value store implementations such as
FloDB~\cite{flodb} would necessarily have commutative interfaces, enabling more
general applicability of Cyclone.
